---
title: "Randall_Boyes_Week2"
author: "Randy Boyes"
date: "09/01/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
```

1. Construct a linear regression of weight as predicted by height, using the
adults (age 18 or greater) from the Howell1 dataset. The heights listed below
were recorded in the !Kung census, but weights were not recorded for these
individuals. Provide predicted weights and 89% compatibility intervals for
each of these

```{r}
data(Howell1)

d <- Howell1
d2 <- d[ d$age >= 18 , ]

flist <- alist(
  weight ~ dnorm( mu , sigma ) ,
  mu <- alpha + height * beta,
  alpha ~ dnorm(0, 10),
  beta ~ dnorm(1, 5),
  sigma ~ dexp(10)
)

m1 <- quap( flist , data=d2 )

```

```{r}
gt::gt(data.frame(precis(m1)))
```

```{r}
sim.weight <- sim(m1, data=list(height=c(140,160,175)))
weight.PI <- apply( sim.weight , 2 , PI , prob=0.89 )
mean <- apply(sim.weight, 2, mean)
gt::gt(data.frame(cbind(c(140, 160, 175), mean, t(weight.PI))))
```


2. From the Howell1 dataset, consider only the people younger than 13 years
old. Estimate the causal association between age and weight. Assume that
age influences weight through two paths. First, age influences height, and
height influences weight. Second, age directly influences weight through age related
changes in muscle growth and body proportions. All of this implies
this causal model (DAG):

Use a linear regression to estimate the total (not just direct) causal effect of
each year of growth on weight. Be sure to carefully consider the priors. Try
using prior predictive simulation to assess what they imply.

```{r}
q2 <- d[d$age < 13, ]

q2l <- list()
q2l$A <- standardize( q2$age )
q2l$W <- standardize( q2$weight )
q2l$H <- standardize( q2$height )

m2 <- quap(
  alist(
  ## A -> W <- H
    W ~ dnorm( mu , sigma ) ,
    mu <- a + bH*H + bA*A ,
    a ~ dnorm( 0 , 0.2 ) ,
    bH ~ dnorm( 0 , 0.5 ) ,
    bA ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 ),
  ## A -> H
    H ~ dnorm( mu_H , sigma_H ),
    mu_H <- aH + bAH*A,
    aH ~ dnorm( 0 , 0.2 ),
    bAH ~ dnorm( 0 , 0.5 ),
    sigma_H ~ dexp( 1 )
) , data = q2l )
```

```{r}
sim_dat <- data.frame( A = seq( from=-2 , to=2 , length.out=30 ) )
s <- sim( m2 , data=sim_dat , vars=c("H","W") )

plot( sim_dat$A , colMeans(s$W) , ylim=c(-2,2) , type="l" ,
xlab="manipulated A" , ylab="counterfactual W" )
shade( apply(s$W,2,PI) , sim_dat$A )
mtext( "Total counterfactual effect of A on W" )


```
In kg per year:

```{r}
(mean(s$W[,30]) - mean(s$W[,1]))/4 * sd(q2$weight)/sd(q2$age)
```


3. Now suppose the causal association between age and weight might be different
for boys and girls. Use a single linear regression, with a categorical
variable for sex, to estimate the total causal effect of age on weight separately
for boys and girls. How do girls and boys differ? Provide one or more posterior
contrasts as a summary.

```{r}
q3 <- d[d$age < 13, ]

q3l <- list()
q3l$A <- standardize( q3$age )
q3l$W <- standardize( q3$weight )
q3l$H <- standardize( q3$height )
q3l$M <- q3$male

m3 <- quap(
  alist(
  ## A -> W <- H
    W ~ dnorm( mu , sigma ) ,
    mu <- aM*M + M*bMH*H + M*bMA*A + a + bH*H + bA*A,
    a ~ dnorm( 0 , 0.2 ) ,
    aM ~ dnorm( 0 , 0.2 ) ,
    bH ~ dnorm( 0 , 0.5 ) ,
    bA ~ dnorm( 0 , 0.5 ) ,
    bMH ~ dnorm( 0 , 0.5 ) ,
    bMA ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 ),
  ## A -> H
    H ~ dnorm( mu_H , sigma_H ),
    mu_H <- aH + M*aMH + bAH*A + M*A*bAHM,
    aH ~ dnorm( 0 , 0.2 ),
    aMH ~ dnorm( 0 , 0.2 ),
    bAH ~ dnorm( 0 , 0.5 ),
    bAHM ~ dnorm( 0 , 0.5 ),
    sigma_H ~ dexp( 1 )
) , data = q3l )
```

```{r}
precis(m3)
```

1. Males are reliably taller than females on average (aM > 0).
2. Males gain more weight for the same amount of height gain (bMH > 0)


